On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-12-04 paper.tex
Rewriting

For $u, v \in \mathbb N$, $\genvar \in \mathbb R^+$, let
\begin{equation}
\label{eq: g def}
g(\genvar, u, v) = \frac 2 \pi \int_{\arcsin \frac v {2\genvar}}^{\arccos \frac u {2\genvar}} \left( \genvar \sin \theta - \frac u 2 \right) \left( \genvar \cos \theta - \frac v 2 \right) \diff \theta.
\end{equation}
Computing the integral in \eqref{eq: g def} and using the identities $\cos^2 \alpha = 1-\sin^2 \alpha$ and $\cos(2\alpha) = 2\cos^2\alpha-1 = 1 - 2\sin^2\alpha$,
\begin{equation}
\label{eq: g fin}
\begin{split}
g(\genvar, u, v) &= \frac 1 {2\pi} \biggl(
\left(\arccos\frac{u}{2\genvar}-\arcsin\frac{v}{2\genvar}\right) u v + 2\genvar^2 + \frac{u^2} 2 + \frac{v^2} 2 \\
& \quad  - u \sqrt{4\genvar^2-v^2} - v \sqrt{4\genvar^2-u^2} \biggr).
\end{split}
\end{equation}

\begin{theorem}
\label{theo: probmax, sq}
Given $\len \in \mathbb R^+$, consider a unit square grid and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles $\tiles = \funt(\len)$ is
\begin{equation}
\label{eq: probmax} %***Change order of the two summands under each square root? here and in the rest of the proof, in accordance with \eqref{eq: square, odd max tiles, lengths}. Or change there
\begin{split}
\probmax(\len) & = g\left(\len, 
\tiles-3, \tiles-3 \right) \quad\ \text{for } \tiles \text{ odd, } \textstyle \len \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = g\left(\len, \tiles-3, \tiles-3 \right) + 2 g\left(\len, \tiles-5, \tiles-1 \right) \\
& \text{for } \tiles \text{ odd, } \textstyle \len \in \left (\frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2,  \frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\len, \tiles-4, \tiles-2 \right) \quad \text{for } \tiles \text{ even, } \textstyle \len \in \left( \frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} 2, \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\len, \tiles-4, \tiles-2 \right) + 2 g\left(\len, \tiles-6, \tiles \right) \\
& \text{for } \tiles \text{ even, } \textstyle \len \in \left( \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2, \frac{\tiles-2}{\sqrt{2}} \right].
\end{split}
\end{equation}
\end{theorem}

@@

For $i,j \in \mathbb Z$, the (square) tile with lower-left coordinate $(i,j)$ is defined as $[i,i+1) \times [j,j+1)$, i.e.~the tile includes its lower and left edges, and tiles are disjoint. The first endpoint of the segment is contained in the tile with lower-left corner $(0,0)$, which will be referred to as reference tile. For an arbitrary value $\tiles$, the segment visits $\tiles$ tiles almost surely if and only if its second endpoint is in a tile with lower-left corner $(i-1,j-1)$, $i,j\geq 1$, $i+j-1=\tiles$, or in a tile symmetrical to this about the line $x=1/2$ (with $i \leq 0$), or about $y=1/2$ (with $j \leq 0$), or about both.

@@

\begin{figure}
\centering%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-1)/2,(\tiles-1)/2)$]{%
\label{fig: probmax_funt_odd_diag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_diag}%
}\\%\hfill%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-3)/2,(\tiles+1)/2)$]{%
\label{fig: probmax_funt_odd_offdiag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_offdiag}%
}%
\caption{Segment orientations and positions of the first endpoint within the reference tile for a given number $\tiles$ of visited tiles. Example for odd $\tiles$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

There is a range of possible values of $\theta$ for which the above condition is satisfied. This range corresponds to the part of the arc with solid line in Figure~\ref{fig: probmax_funt_odd}. For convenience, the angle $\theta$ is defined clockwise from the downward direction. For each $\theta$ in the allowed range, the valid positions for the first endpoint of the (shifted) segment are in a rectangle contained in the reference tile. The rectangle associated with an example $\theta$ is shown in the figure, shaded. The area occupied by this rectangle is the probability $\probmax_{i-1,j-1}(\len,\theta)$ that the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ conditioned on $\theta$. The desired probability $\probmax(\len)$ will be obtained by averaging $\probmax_{i-1,j-1}(\len, \theta)$ over $\theta$, summing over all $i, j \geq 1$ such that the tile with lower-left corner $(i-1, j-1)$ can be reached from the reference tile, and then using symmetry to take into account the tiles with $i \leq 0$ or $j \leq 0$.

The situation is different depending on whether $\tiles = \funt(\len)$ is odd or even. Consider $\tiles$ odd first. Under this assumption, the range of lengths for which $\funt(\len)$ equals $\tiles$ is given by \eqref{eq: square, odd max tiles, lengths}. For any length in this interval, there are some positions within the reference tile from which the tile with lower-left coordinate $((\tiles-1)/2, (\tiles-1)/2)$ can be reached. Equivalently, the shifted segment with one endpoint at $((\tiles-1)/2, (\tiles-1)/2)$ can reach the reference tile. This is illustrated in Figure~\ref{fig: probmax_funt_odd_diag}. In addition, if 
\begin{equation}
\label{eq: probmax len odd2}
\len > \frac{\sqrt{(\tiles-5)^2 + (\tiles-1)^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $((\tiles-3)/2, (\tiles+1)/2)$ and $((\tiles+1)/2, (\tiles-3)/2)$ are also reachable from the reference tile; or equivalently the shifted segment can reach the reference tile, as shown in Figure~\ref{fig: probmax_funt_odd_offdiag}. This can only happen for $\tiles \geq 5$ (for $\tiles=3$ the inequality \eqref{eq: probmax len odd2} cannot be satisfied for lengths in the range given by \eqref{eq: square, odd max tiles, lengths}). In this case, by symmetry, each of those two tiles contributes the same probability. Other tiles with $i,j\geq 1$, $i+j-1 = \tiles$ do not need to be considered, because they cannot be reached with lengths satisfying \eqref{eq: square, odd max tiles, lengths}.

@@

The above two paragraphs show that, among all possible tiles with lower-left corner $(i-1,j-1)$, $i,j\geq 1$, those whose centers lie on the lines $x=1/2$ or $y=1/2$ (i.e.~those with $i=1$ of $j=1$) are never reachable from the reference tile, in the sense that a segment that touches the maximum number of tiles cannot have its second endpoint in those tiles. This implies that $\probmax(\len)$ can be computed by considering only the tiles with $i,j\geq 1$ (actually $i,j \geq 2$), and then multiplying by $4$ to include the symmetrical tiles with respect to $x=1/2$ and $y=1/2$. As an example, Figure~\ref{fig: probmax_possible_tiles} shows the reference tile (solid line) and the tiles that can contain the second endpoint (dotted lines) for $\funt(\len) = 5$.

\begin{figure}
\centering%
\includegraphics[width=.52\textwidth]{probmax_possible_tiles}%
\caption{Tiles that can contain the second endpoint of a segment that visits the maximum number of tiles $\funt(\len)$ in a square grid (example for $\funt(\len)=5$)%
}%
\label{fig: probmax_possible_tiles}%
\end{figure}%

According to the preceding analysis, two cases need to be distinguished for $\tiles = \funt(\len)$ odd, depending on whether \eqref{eq: probmax len odd2} is satisfied or not; and similarly two cases exist for $\tiles = \funt(\len)$ even, depending on whether \eqref{eq: probmax len even2} holds or not. These are, respectively, the four cases in \eqref{eq: probmax}.

In the first case indicated in \eqref{eq: probmax}, as previously argued, $\tiles$ is odd and only the tile with lower-left coordinates $((\tiles-1)/2, (\tiles-1)/2)$ is reachable. For a given $\theta$, the conditional probability $\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta)$ is the area of the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_diag}:
\begin{equation}
\label{eq: probmax: probmax theta odd}
\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta) = \left(\len\sin\theta - \frac{\tiles-3}2\right) \left(\len\cos\theta - \frac{\tiles-3}2\right).
\end{equation}
The range of allowed values for $\theta$, as can be seen from the figure, is $(\theta_0, \theta_1)$ with
\begin{align}
\label{eq: odd theta 0}
\theta_0 & = \arcsin\frac{\tiles-3}{2\len}, \\
\label{eq: odd theta 1}
\theta_1 & = \arccos\frac{\tiles-3}{2\len}.
\end{align}
Averaging \eqref{eq: probmax: probmax theta odd} over $\theta$, multiplying by $4$ to include the symmetrical tiles in other quadrants and using \eqref{eq: g def} gives $\probmax(\len)$ as
\begin{equation}
\label{eq: probmax: probmax odd}
\begin{split}
\probmax(\len) & = 4 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta)}{2\pi} \diff \theta \\
& = \frac 2 \pi \int_{\arcsin\frac{\tiles-3}{2\len}}^{\arccos\frac{\tiles-3}{2\len}} \left( \len\sin\theta - \frac{\tiles-3} 2 \right)  \left( \len\cos\theta - \frac{\tiles-3} 2 \right) \diff \theta \\
& = g\left( \len, \tiles-3, \tiles-3 \right ),
\end{split}
\end{equation}
in accordance with \eqref{eq: probmax}.

>>>>

For $u, v \in \mathbb N$, $\genvar \in \mathbb R^+$, let
\begin{equation}
\label{eq: g def}
g(\genvar, u, v) = \frac 2 \pi \int_{\arcsin \frac u {2\genvar}}^{\arccos \frac v {2\genvar}} \left( \genvar \sin \theta - \frac u 2 \right) \left( \genvar \cos \theta - \frac v 2 \right) \diff \theta.
\end{equation}
Computing the integral in \eqref{eq: g def} and using the identities $\cos^2 \alpha = 1-\sin^2 \alpha$ and $\cos(2\alpha) = 2\cos^2\alpha-1 = 1 - 2\sin^2\alpha$,
\begin{equation}
\label{eq: g fin}
\begin{split}
g(\genvar, u, v) &= \frac 1 {2\pi} \biggl(
\left(\arccos\frac{v}{2\genvar}-\arcsin\frac{u}{2\genvar}\right) u v + 2\genvar^2 + \frac{u^2} 2 + \frac{v^2} 2 \\
& \quad  - u \sqrt{4\genvar^2-v^2} - v \sqrt{4\genvar^2-u^2} \biggr).
\end{split}
\end{equation}

\begin{theorem}
\label{theo: probmax, sq}
Given $\len \in \mathbb R^+$, consider a unit square grid and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles $\tiles = \funt(\len)$ is
\begin{equation}
\label{eq: probmax} %***Change order of the two summands under each square root here and in the rest of the proof, in accordance with \eqref{eq: square, odd max tiles, lengths}. (Or change there, but I think that is less preferable because it would correspond to using the y coordinate, then the x coordinate from the figure)
\begin{split}
\probmax(\len) & = g\left(\len, 
\tiles-3, \tiles-3 \right) \quad\ \text{for } \tiles \text{ odd, } \textstyle \len \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = g\left(\len, \tiles-3, \tiles-3 \right) + 2 g\left(\len, \tiles-5, \tiles-1 \right) \\
& \text{for } \tiles \text{ odd, } \textstyle \len \in \left (\frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2,  \frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\len, \tiles-4, \tiles-2 \right) \quad \text{for } \tiles \text{ even, } \textstyle \len \in \left( \frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} 2, \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\len, \tiles-4, \tiles-2 \right) + 2 g\left(\len, \tiles-6, \tiles \right) \\
& \text{for } \tiles \text{ even, } \textstyle \len \in \left( \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2, \frac{\tiles-2}{\sqrt{2}} \right].
\end{split}
\end{equation}
\end{theorem}

@@

For $i,j \in \mathbb Z$, the (square) tile with lower-left coordinate $(i,j)$ is defined as $[i,i+1) \times [j,j+1)$, i.e.~including its lower and left edges. The first endpoint of the segment is contained in the tile with lower-left corner $(0,0)$, which will be referred to as reference tile. For an arbitrary value $\tiles$, the segment visits $\tiles$ tiles almost surely if its second endpoint is in a tile with lower-left corner $(i-1,j-1)$ such that $i,j\geq 1$ and $i+j-1=\tiles$, or in a tile symmetrical to this about the line $x=1/2$ (with $i \leq 0$), or about $y=1/2$ (with $j \leq 0$), or about both.

@@

\begin{figure}
\centering%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-1)/2,(\tiles-1)/2)$]{%
\label{fig: probmax_funt_odd_diag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_diag}%
}\\%\hfill%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-3)/2,(\tiles+1)/2)$]{%
\label{fig: probmax_funt_odd_offdiag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_offdiag}%
}%
\caption{Segment orientations and positions of the first endpoint within the reference tile for a given $\tiles = \funt(\len)$, with $\tiles$ odd. Example for $\tiles = 7$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

There is a range of possible values of $\theta$ for which the above condition is satisfied. This range corresponds to the part of the arc with solid line in Figure~\ref{fig: probmax_funt_odd}. For convenience, the angle $\theta$ is defined clockwise from the downward direction. For each $\theta$ in the allowed range, the valid positions for the first endpoint of the (shifted) segment are in a rectangle contained in the reference tile. The rectangle associated with an example $\theta$ is shown in the figure, shaded. The area of this rectangle is the probability $\probmax_{i-1,j-1}(\len,\theta)$ that the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ conditioned on $\theta$. The desired probability $\probmax(\len)$ will be obtained by averaging $\probmax_{i-1,j-1}(\len, \theta)$ over $\theta$, summing over all $i, j \geq 1$ such that the tile with lower-left corner $(i-1, j-1)$ can be reached from the reference tile, and then using symmetry to take into account the tiles with $i \leq 0$ or $j \leq 0$.

The situation is different depending on whether $\tiles = \funt(\len)$ is odd or even. Consider $\tiles$ odd first. Under this assumption, the range of lengths for which $\funt(\len)$ equals $\tiles$ is given by \eqref{eq: square, odd max tiles, lengths}. For any length in this interval, there are some positions within the reference tile from which the tile with lower-left coordinate $((\tiles-1)/2, (\tiles-1)/2)$ can be reached. Equivalently, the shifted segment with one endpoint at $((\tiles-1)/2, (\tiles-1)/2)$ can reach the reference tile. This is illustrated in Figure~\ref{fig: probmax_funt_odd_diag}. In addition, if 
\begin{equation}
\label{eq: probmax len odd2}
\len > \frac{\sqrt{(\tiles-5)^2 + (\tiles-1)^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $((\tiles-3)/2, (\tiles+1)/2)$ and $((\tiles+1)/2, (\tiles-3)/2)$ are also reachable from the reference tile; or equivalently the shifted segment can reach the reference tile, as shown in Figure~\ref{fig: probmax_funt_odd_offdiag}. This can only happen for $\tiles \geq 5$ (for $\tiles=3$ the inequality \eqref{eq: probmax len odd2} cannot be satisfied for lengths in the range given by \eqref{eq: square, odd max tiles, lengths}). By symmetry, each of those two tiles contributes the same probability. Other tiles with $i,j\geq 1$, $i+j-1 = \tiles$ do not need to be considered, because they cannot be reached with lengths satisfying \eqref{eq: square, odd max tiles, lengths}.

@@

From the above two paragraphs it stems that, among all possible tiles with lower-left corner $(i-1,j-1)$, $i,j\geq 1$, those whose centers lie on the lines $x=1/2$ or $y=1/2$ (i.e.~those with $i=1$ of $j=1$) are never reachable from the reference tile, in the sense that a segment that touches the maximum number of tiles cannot have its second endpoint in those tiles. This implies that $\probmax(\len)$ can be computed by considering only the tiles with $i,j\geq 1$ (actually $i,j \geq 2$), and then multiplying by $4$ to include the symmetrical tiles with respect to the lines $x=1/2$ and $y=1/2$. As an example, Figure~\ref{fig: probmax_possible_tiles} shows the reference tile (solid line) and the tiles that can contain the second endpoint (dotted lines) for $\funt(\len) = 5$.

\begin{figure}
\centering%
\includegraphics[width=.52\textwidth]{probmax_possible_tiles}%
\caption{Tiles that can contain the second endpoint of a segment that visits the maximum number of tiles $\tiles= \funt(\len)$, with $\tiles$ odd. Example for $\tiles = 5$%
}%
\label{fig: probmax_possible_tiles}%
\end{figure}%

According to the preceding analysis, two cases need to be distinguished for $\tiles = \funt(\len)$ odd, depending on whether \eqref{eq: probmax len odd2} is satisfied or not; and similarly two cases exist for $\tiles = \funt(\len)$ even, depending on whether \eqref{eq: probmax len even2} holds or not. These are the four cases in \eqref{eq: probmax}.

In the first case indicated in \eqref{eq: probmax}, as previously argued, $\tiles$ is odd and only the tile with lower-left coordinates $((\tiles-1)/2, (\tiles-1)/2)$ is reachable. For a given $\theta$, the conditional probability $\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta)$ is the area of the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_diag}:
\begin{equation}
\label{eq: probmax: probmax theta odd}
\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta) = \left(\len\sin\theta - \frac{\tiles-3}2\right) \left(\len\cos\theta - \frac{\tiles-3}2\right).
\end{equation}
The range of allowed values for $\theta$, as can be deduced from the figure, is $(\theta_0, \theta_1)$ with
\begin{align}
\label{eq: odd theta 0}
\theta_0 & = \arcsin\frac{\tiles-3}{2\len}, \\
\label{eq: odd theta 1}
\theta_1 & = \arccos\frac{\tiles-3}{2\len}.
\end{align}
Averaging \eqref{eq: probmax: probmax theta odd} over $\theta$, multiplying by $4$ to include the symmetrical tiles in other quadrants and using \eqref{eq: g def} gives $\probmax(\len)$ as
\begin{equation}
\label{eq: probmax: probmax odd}
\begin{split}
\probmax(\len) & = 4 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta)}{2\pi} \diff \theta \\
& = \frac 2 \pi \int_{\arcsin\frac{\tiles-3}{2\len}}^{\arccos\frac{\tiles-3}{2\len}} \left( \len\sin\theta - \frac{\tiles-3} 2 \right)  \left( \len\cos\theta - \frac{\tiles-3} 2 \right) \diff \theta \\
& = g\left( \len, \tiles-3, \tiles-3 \right ),
\end{split}
\end{equation}
in accordance with the first expression in \eqref{eq: probmax}.
